A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-16 Manuscript/OldVsns/Manuscript.Rmd

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers
% https://github.com/rstudio/rmarkdown/issues/337
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

% https://github.com/rstudio/rmarkdown/pull/252
\usepackage{titling}
\setlength{\droptitle}{-2em}

\pretitle{\vspace{\droptitle}\centering\huge}
\posttitle{\par}

\preauthor{\centering\large\emph}
\postauthor{\par}

\predate{\centering\large\emph}
\postdate{\par}

\date{}

\begin{document}

\section{Pagel's Lambda Estimates are Often
Inaccurate}\label{pagels-lambda-estimates-are-often-inaccurate}

@@

\textbf{Keywords}: Pagel's lambda, phylogenetic signal \hfill\break

\textbf{Short Title}: Inaccuracies in Pagel's Lambda \hfill\break

\section{Abstract}\label{abstract}

\{conclusion holds: interpreting the regression is not appreciably
different (in terms of slopes and f values)\}

@@

\section{Introduction}\label{introduction}

Investigating macroevolutionary patterns requires a phylogenetic
approach as species are non-independent by nature of their shared
ancestry. Since the first appropriate method was introduced by
Felsenstein (phylogenetic independent contrasts; Felsenstein (1985)),
dozens of other methods have been developed and applied to increasingly
complex questions in macroevolutionary biology (e.g. Grafen (1989);
Harvey and Pagel (1991); Rohlf (2001); ({\textbf{???}})). A particularly
useful measure in this field is phylogenetic signal, a quantification of
a trait's correspondence with the clade's phylogenic history.
Understanding the degree of phylogenetic signal present in a dataset is
paramount and identifies the mode under which a trait has evolved; high
measures of phylogenetic signal indicate a Brownian motion process,
whereas lower levels of phylogenetic signal indicate natural selection
or some other evolutionary force has influenced the traits evolutionary
history. \hfill\break

Several approaches to quantify phylogenetic signal exist. For continuous
data, the most common parameters used in the literature include Pagel's
lambda ({\textbf{???}}) and Blomberg's kappa ({\textbf{???}}). Pagel's
lambda has the advantage of being nested in the likelihood framework and
thus has also been utilized to simultaneously estimate and account for
phylogenetic signal while doing phylogenetic regressions or ANOVAs. At
the development of these estimation methods, clear recommendations on
the use and interpretation of this parameter were made based on data
simulations and error evaluations (Revell 2010). However, the accuracy
of the lambda estiamtion methods have not been fully evaluated, and
consequently many studies have substantially misused these methods.
\hfill\break

An earlier study ({\textbf{???}}), breifly addressed this topic by
showing how uninformative smaller phylogenies are when estimating
various evolutionary parameters, such as Pagel's lambda. However, it
remains unknown the degree to which lambda estimates accurately
represent degree of phylogenetic signal, and thus we fully explicate the
extent to which Pagel's lambda estimation methods can and have been
misused in recent macroevolutionary studies. Here we take a
comprehensive approach to demonstrate the scenarios under which
estimated Pagel's lambdas accurately reflect known lambdas. We then
demonstrate the effect of these, at times, dubious estimation methods on
significance testing when used in a PGLS framework. Finally, we perform
a meta-analysis of research published in 2019 employing Pagel's lambda
estimation methods to demonstrate the uses and misuses of this method in
the literature. This investigation provides evidence for the
circumstances under which estimating Pagel's lambda can be informative,
and which interpretations or uses are misleading with the hope of moving
the field towards more appropriate employment of phylogenetic
comparative methods.

\section{Methods and Results}\label{methods-and-results}

\subsection{\texorpdfstring{\emph{Simulated
trait}}{Simulated trait}}\label{simulated-trait}

@@

To visualize the accuracy of the lambda estimation methods, we first
plotted known lambdas (input lambdas) against the estimated lambdas
(Figure 1). This plot demonstrates the rampant inaccuracy of estimating
lambdas on phylogenies with fewer than 200 tips, as the spread of data
in the upper panels in Figure 1 is remarkably wide. We also see that the
widest spread of estimated lambdas is observed near the center of each
plot corresponding to intermediate values of known lambda. Lastly, we
see a slight tendency for the PGLS estimation methods to underestimate
lambda, especially for known lambda values below 0.5 as can be seen by
the numerous data points along the x axis for the smaller phylogenies
analyzed (n tips \textless{} 200). \hfill\break

{[}insert Figure 1 here{]}

\subsection{\texorpdfstring{\emph{Simulated ANOVA and
Regressions}}{Simulated ANOVA and Regressions}}\label{simulated-anova-and-regressions}

@@

{[}insert Figure 2 here{]} \hfill\break

@@

Scaling and trait generation procedures for all simulation methods were
repeated with symmetrical and ladder phylogenies of variable size.
Results generated using these variable phylogeny shapes were consistent
with the pure-birth phylogeny results presented above and can be found
in the Supplemental Materials. All analyses were performed in R v3.6.0
({\textbf{???}}) using the packages \texttt{geiger} ({\textbf{???}}) and
\texttt{caper} ({\textbf{???}}), and the corresponding scripts can be
found in the Supplemental Materials.

\subsection{\texorpdfstring{\emph{Meta-Analysis of Empirical
Results}}{Meta-Analysis of Empirical Results}}\label{meta-analysis-of-empirical-results}

To understand the extent of this problematic estimation method in
application, we performed a meta-analysis of studies published in 2019
that cited Pagel's 1997 manuscript ({\textbf{???}}). The list of
manuscripts was compiled through Google Scholar on Jan 23, 2020 and
totaled 341 manuscripts. For each study, we extracted any published
lambda estimates, along with the size of the phylogeny used in the
analysis. We also noted whether authors reported confidence intervals,
significance tests assessing difference of the lambda estimate from 0 or
1, and whether authors interpreted biological meaning from the magnitude
of the estimated lambda. For studies that reported more than one lambda
estimate, we also noted if the authors compared the lambdas against one
another, and whether that was accompanied with an appropriate
statistical test between the estimated lambda values. \hfill\break

We found 182 manuscripts from 2019 that estimated and reported Pagel's
lambda values using PGLS methods. These papers averaged 8.527 lambda
values, ranging from a single lambda estimate up to 71 estimated
lambdas. Almost exactly half of the published lambda estimates were
either below 0.05 (25.32\%) or above 0.9 (24.74\%; Figure 3). 73.32\% of
the published lambdas were estimated using phylogenies with fewer than
200 tips, and 348 lambda estimates (8.57\% of all published estimates)
came from phylogenies with fewer than 30 tips. \hfill\break

{[}insert Figure 3 here{]} \hfill\break

Many of the reviewed manuscripts liberally interpreted the magnitude of
the estimate lambda, using phrases such as ``strong'' or ``weak''
phylogentic signal when statistically, all that was clear was a
difference between the estimated lambda and 0 or 1 respectively. We
estimated that about 20.49\% of the manuscripts revealed some sort of
biological interpretation of the magnitude of estimated phylogenetic
signal that overreached the statistical findings. We also identified
seven manuscripts as having inappropriately interpreted differences in
lambda values, indicating that some traits had stronger or weaker signal
than other traits without the appropriate statistical tests.
\hfill\break

As is evidenced by macroevolutionary papers published in 2019 papers,
Pagel's lambda estimation methods are often misused and
over-interpretted. Despite the urging of Boettiger and colleagues to
publish confidence intervals with all lambda parameter estimates, only
18\% of papers published in 2019 do so.

\section{Discussion}\label{discussion}

{[}This part is obviously not written yet{]} \hfill\break

General conclusions :Using the estimated lambda values from pgls are not
useful. The questions of whether or not signal exists is appropriate,
but inferring more from lambda \emph{magnitude} is inappropriate.
\hfill\break

@@

\section{References}\label{references}

@@

\hypertarget{refs}{}
\hypertarget{ref-Felsenstein1985}{}
Felsenstein, J. 1985. Phylogenies and the comparative method. American
Naturalist 125:1--15.

\hypertarget{ref-Grafen1989}{}
Grafen, A. 1989. The phylogenetic regression. Philosophical Transactions
of the Royal Society of London B, Biological Sciences 326:119--157.

\hypertarget{ref-HarveyPagel1991}{}
Harvey, P. H., and M. D. Pagel. 1991. The comparative method in
evolutionary biology. Oxford University Press, Oxford.

\hypertarget{ref-Revell2010}{}
Revell, L. J. 2010. Phylogenetic signal and linear regression on species
data. Methods in Ecology and Evolution 1:319--329.

\hypertarget{ref-Rohlf2001}{}
Rohlf, F. J. 2001. Comparative methods for the analysis of continuous
variables: Geometric interpretations. Evolution 55:2143--2160.

@@

\section{Figure Legends}\label{figure-legends}

@@

\includegraphics[width=0.95\linewidth]{Fig1}

@@

\includegraphics[width=0.95\linewidth]{Fig2}

@@

\includegraphics[width=0.95\linewidth]{Fig3}

@@

\end{document}

>>>>

---
title:

output: 
  pdf_document:
    fig_caption: yes
    number_sections: no  #note: change to 'no' for most journals
    keep_tex: true
    pandoc_args:

csl: evolution.csl
bibliography: bibliography.bib

header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers

---
# Pagel's Lambda Estimates are Often Inaccurate

@@

**Keywords**: Pagel's lambda, phylogenetic signal \hfill\break

**Short Title**: Inaccuracies in Pagel's Lambda \hfill\break

```{r setup, include=FALSE}
  library(knitr)
  knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

{conclusion holds: interpreting the regression is not appreciably different (in terms of slopes and f values)}

@@

# Introduction
Investigating macroevolutionary patterns requires a phylogenetic approach as species are non-independent by nature of their shared ancestry. Since the first appropriate method was introduced by Felsenstein (phylogenetic independent contrasts; @Felsenstein1985), dozens of other methods have been developed and applied to increasingly complex questions in macroevolutionary biology (e.g. @Grafen1989; @HarveyPagel1991; @Rohlf2001; @AdamsNason2019). A particularly useful measure in this field is phylogenetic signal, a quantification of a trait's correspondence with the clade's phylogenic history. Understanding the degree of phylogenetic signal present in a dataset is paramount and identifies the mode under which a trait has evolved; high measures of phylogenetic signal indicate a Brownian motion process, whereas lower levels of phylogenetic signal indicate natural selection or some other evolutionary force has influenced the traits evolutionary history. \hfill\break

Several approaches to quantify phylogenetic signal exist. For continuous data, the most common parameters used in the literature include Pagel's lambda [@Pagel1997] and Blomberg's kappa [@Blomberg1900]. Pagel's lambda has the advantage of being nested in the likelihood framework and thus has also been utilized to simultaneously estimate and account for phylogenetic signal while doing phylogenetic regressions or ANOVAs. At the development of these estimation methods, clear recommendations on the use and interpretation of this parameter were made based on data simulations and error evaluations [@Revell2010]. However, the accuracy of the lambda estiamtion methods have not been fully evaluated, and consequently many studies have substantially misused these methods.  \hfill\break

An earlier study [@BoettigerEtAl2012], breifly addressed this topic by showing how uninformative smaller phylogenies are when  estimating various evolutionary parameters, such as Pagel's lambda. However, it remains unknown the degree to which lambda estimates accurately represent degree of phylogenetic signal, and thus we fully explicate the extent to which Pagel's lambda estimation methods can and have been misused in recent macroevolutionary studies. Here we take a comprehensive approach to demonstrate the scenarios under which estimated Pagel's lambdas accurately reflect known lambdas. We then demonstrate the effect of these, at times, dubious estimation methods on significance testing when used in a PGLS framework. Finally, we perform a meta-analysis of research published in 2019 employing Pagel's lambda estimation methods to demonstrate the uses and misuses of this method in the literature. This investigation provides evidence for the circumstances under which estimating Pagel's lambda can be informative, and which interpretations or uses are misleading with the hope of moving the field towards more appropriate employment of phylogenetic comparative methods.

# Methods and Results

## *Simulated trait*

@@

To visualize the accuracy of the lambda estimation methods, we first plotted known lambdas (input lambdas) against the estimated lambdas (Figure 1). This plot demonstrates the rampant inaccuracy of estimating lambdas on phylogenies with fewer than 200 tips, as the spread of data in the upper panels in Figure 1 is remarkably wide. We also see that the widest spread of estimated lambdas is observed near the center of each plot corresponding to intermediate values of known lambda. Lastly, we see a slight tendency for the PGLS estimation methods to underestimate lambda, especially for known lambda values below 0.5 as can be seen by the numerous data points along the x axis for the smaller phylogenies analyzed (n tips < 200). \hfill\break

[insert Figure 1 here]

## *Simulated ANOVA and Regressions*

@@

[insert Figure 2 here] \hfill\break

@@

Scaling and trait generation procedures for all simulation methods were repeated with symmetrical and ladder phylogenies of variable size. Results generated using these variable phylogeny shapes were consistent with the pure-birth phylogeny results presented above and can be found in the Supplemental Materials. All analyses were performed in R v3.6.0 [@RCoreTeam] using the packages `geiger` [@HarmonEtAl2019] and `caper` [@Orme2011], and the corresponding scripts can be found in the Supplemental Materials.

## *Meta-Analysis of Empirical Results*

To understand the extent of this problematic estimation method in application, we performed a meta-analysis of studies published in 2019 that cited Pagel's 1997 manuscript [@Pagel1997]. The list of manuscripts was compiled through Google Scholar on Jan 23, 2020 and totaled 341 manuscripts. For each study, we extracted any published lambda estimates, along with the size of the phylogeny used in the analysis. We also noted whether authors reported confidence intervals, significance tests assessing difference of the lambda estimate from 0 or 1, and whether authors interpreted biological meaning from the magnitude of the estimated lambda. For studies that reported more than one lambda estimate, we also noted if the authors compared the lambdas against one another, and whether that was accompanied with an appropriate statistical test between the estimated lambda values. \hfill\break

We found 182 manuscripts from 2019 that estimated and reported Pagel's lambda values using PGLS methods. These papers averaged 8.527 lambda values, ranging from a single lambda estimate up to 71 estimated lambdas. Almost exactly half of the published lambda estimates were either below 0.05 (25.32%) or above 0.9 (24.74%; Figure 3). 73.32% of the published lambdas were estimated using phylogenies with fewer than 200 tips, and 348 lambda estimates (8.57% of all published estimates) came from phylogenies with fewer than 30 tips.  \hfill\break

[insert Figure 3 here] \hfill\break

Many of the reviewed manuscripts liberally interpreted the magnitude of the estimate lambda, using phrases such as "strong" or "weak" phylogentic signal when statistically, all that was clear was a difference between the estimated lambda and 0 or 1 respectively. We estimated that about 20.49% of the manuscripts revealed some sort of biological interpretation of the magnitude of estimated phylogenetic signal that overreached the statistical findings. We also identified seven manuscripts as having inappropriately interpreted differences in lambda values, indicating that some traits had stronger or weaker signal than other traits without the appropriate statistical tests. \hfill\break

As is evidenced by macroevolutionary papers published in 2019 papers, Pagel's lambda estimation methods are often misused and over-interpretted. Despite the urging of Boettiger and colleagues to publish confidence intervals with all lambda parameter estimates, only 18% of papers published in 2019 do so.

# Discussion

[This part is obviously not written yet] \hfill\break

General conclusions :Using the estimated lambda values from pgls are not useful. The questions of whether or not signal exists is appropriate, but inferring more from lambda *magnitude* is inappropriate. \hfill\break

@@

# References

@@

<div id="refs"></div>

@@

# Figure Legends

@@

```{r, echo = FALSE, out.width="95%"}
include_graphics("Fig1.png")

```

@@

\newpage
```{r, echo = FALSE, out.width="95%"}
include_graphics("Fig2.png")

```

@@

```{r, echo = FALSE, out.width="95%"}
include_graphics("Fig3.png")

```

@@

### 2020-07-16 Manuscript/OldVsns/Manuscript.tex

\section{Pagel's Lambda Estimates are Often
Inaccurate}\label{pagels-lambda-estimates-are-often-inaccurate}

@@

\section{Introduction}\label{introduction}

Investigating macroevolutionary patterns requires a phylogenetic
approach as species are non-independent by nature of their shared
ancestry. Since the first appropriate method was introduced by
Felsenstein (phylogenetic independent contrasts; Felsenstein (1985)),
dozens of other methods have been developed and applied to increasingly
complex questions in macroevolutionary biology (e.g. Grafen (1989);
Harvey and Pagel (1991); Rohlf (2001); ({\textbf{???}})). A particularly
useful measure in this field is phylogenetic signal, a quantification of
a trait's correspondence with the clade's phylogenic history.
Understanding the degree of phylogenetic signal present in a dataset is
paramount and identifies the mode under which a trait has evolved; high
measures of phylogenetic signal indicate a Brownian motion process,
whereas lower levels of phylogenetic signal indicate natural selection
or some other evolutionary force has influenced the traits evolutionary
history. \hfill\break

@@

\section{Methods and Results}\label{methods-and-results}

\subsection{\texorpdfstring{\emph{Simulated
trait}}{Simulated trait}}\label{simulated-trait}

@@

\subsection{\texorpdfstring{\emph{Meta-Analysis of Empirical
Results}}{Meta-Analysis of Empirical Results}}\label{meta-analysis-of-empirical-results}

@@

\hypertarget{refs}{}
\hypertarget{ref-Felsenstein1985}{}
Felsenstein, J. 1985. Phylogenies and the comparative method. American
Naturalist 125:1--15.

\hypertarget{ref-Grafen1989}{}
Grafen, A. 1989. The phylogenetic regression. Philosophical Transactions
of the Royal Society of London B, Biological Sciences 326:119--157.

\hypertarget{ref-HarveyPagel1991}{}
Harvey, P. H., and M. D. Pagel. 1991. The comparative method in
evolutionary biology. Oxford University Press, Oxford.

\hypertarget{ref-Revell2010}{}
Revell, L. J. 2010. Phylogenetic signal and linear regression on species
data. Methods in Ecology and Evolution 1:319--329.

\hypertarget{ref-Rohlf2001}{}
Rohlf, F. J. 2001. Comparative methods for the analysis of continuous
variables: Geometric interpretations. Evolution 55:2143--2160.

>>>>

\hypertarget{pagels-lambda-estimates-are-often-inaccurate}{%
\section{Pagel's Lambda Estimates are Often
Inaccurate}\label{pagels-lambda-estimates-are-often-inaccurate}}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Investigating macroevolutionary patterns requires a phylogenetic
approach as species are non-independent by nature of their shared
ancestry. Since the first appropriate method was introduced by
Felsenstein (phylogenetic independent contrasts; Felsenstein (1985)),
dozens of other methods have been developed and applied to increasingly
complex questions in macroevolutionary biology (e.g.~Grafen (1989);
Harvey and Pagel (1991); Rohlf (2001); ({\textbf{???}})). A particularly
useful measure in this field is phylogenetic signal, a quantification of
a trait's correspondence with the clade's phylogenic history.
Understanding the degree of phylogenetic signal present in a dataset is
paramount and identifies the mode under which a trait has evolved; high
measures of phylogenetic signal indicate a Brownian motion process,
whereas lower levels of phylogenetic signal indicate natural selection
or some other evolutionary force has influenced the traits evolutionary
history. \hfill\break

@@

\hypertarget{methods-and-results}{%
\section{Methods and Results}\label{methods-and-results}}

\hypertarget{simulated-trait}{%
\subsection{\texorpdfstring{\emph{Simulated
trait}}{Simulated trait}}\label{simulated-trait}}

@@

\hypertarget{meta-analysis-of-empirical-results}{%
\subsection{\texorpdfstring{\emph{Meta-Analysis of Empirical
Results}}{Meta-Analysis of Empirical Results}}\label{meta-analysis-of-empirical-results}}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Felsenstein1985}{}%
Felsenstein, J. 1985. Phylogenies and the comparative method. American
Naturalist 125:1--15.

\leavevmode\hypertarget{ref-Grafen1989}{}%
Grafen, A. 1989. The phylogenetic regression. Philosophical Transactions
of the Royal Society of London B, Biological Sciences 326:119--157.

\leavevmode\hypertarget{ref-HarveyPagel1991}{}%
Harvey, P. H., and M. D. Pagel. 1991. The comparative method in
evolutionary biology. Oxford University Press, Oxford.

\leavevmode\hypertarget{ref-Revell2010}{}%
Revell, L. J. 2010. Phylogenetic signal and linear regression on species
data. Methods in Ecology and Evolution 1:319--329.

\leavevmode\hypertarget{ref-Rohlf2001}{}%
Rohlf, F. J. 2001. Comparative methods for the analysis of continuous
variables: Geometric interpretations. Evolution 55:2143--2160.
